### 2013-01-24 patchy-selection-paper.tex
/* Patchy Selection paper */

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{natbib}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{latexml} % for \iflatexml

@@

@@

Convergent adaptation can also occur within species, with individuals within a species 
adapted to the same environment in parallel through different large effect changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes REFs.
All else being equal the evolution of convergent phenotypes within a species convergence is more likely with a higher mutational input, i.e. when the mutation target and population sizes are larger \citep{}. 
Convergence is also more likely when standing variation is present for the phenotype, 
as multiple pre-existing alleles can spread from low frequency, 
if our now adaptive alleles were previously not too deleterious \citep{Orr,Hermission}. 
When multiple rare changes sweep up from low frequency in parallel within a population this has been termed a soft sweep.\\

The geographic distribution of populations can also shape the probability of parallel mutation within a species.
A geographically wide spread species is more likely to adapt by multiple mutations im parallel if dispersal 
is geographically limited, as populations will adapt via new mutations rather than waiting for migration \citep{RalphCoop}. 
Convergence is also more likely if geographically separated populations adapting to similar environments 
via alleles which are maladapted in the environmentals separating them. 
Such adverse conditions can strongly spatially restrict the spread of locally adapted alleles \citep{lenormand} 
and so might encourage local adaptation by parallel mutation rather than migration.\\

@@

In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
where the selection pressure was constant across the entire environment, 
and that a single mutational change was sufficient to adapt populations to the novel environment. 
We assumed that there was no standing variation for the adaptive allele, such that parallel mutation must be due to multiple mutations after the environmental shift. In that setting we found that there was a characteristic geographic scale, over which we would expect multiple instances of our adaptive allele to have arisen in parallel. This characteristic length could be expressed in terms of a simple compound parameter determined by our parameters of interest.
Recall questions of origin of adaptations

References on parallel adapation, standing variation, etc.:
  coop,
  pennings and hermisson.
Note pennings and hermisson found standing deleterious variation was very important.

recall peromyscus examples with patchy environments

Summarize previous paper

Question assumptions of no standing variation and homogenous selection

@@

It takes only a single mutational change to create an allele ($B$) that adapts an individual to the
poor habitat type. This mutational change occurs at a low rate of
$\mu$ per chromosome per generation. We assume that this initially rare new
allele has fitness $1+s_b$ relative to the unmutated type ($b$) in the ``poor'' habitat patches,
and assume that it has fitness $1+s_m$ when rare in the intervening areas, with
$s_m<0$. To simplify matters we assume that the disadvantage $s_m$ ensures that on the relevant timescale,
the allele is very unlikely to fix locally in the regions where it is at a disadvantage.

@@

As only a single mutation is required to adapt individuals to the
poor habitat patch, subsequent mutations that arise after an allele
becomes established at its equilibrium frequency on the patch gain no selective
benefit. Similarly an allele introduced into a patch by migration will
not be favored by selection to spread, if the patch has already been
colonized. Thus mutations selective exclude each other from
patches, over short-time scales, and they will only slowly filter
together over longer time scales by drift.

\paragraph{Previous work} We will make use of a number previous
results from the literature. \citet{slatkin1973geneflow}, working in
one dimensional deterministic framework, showed that if the
physical width of the patch is less than $2 \tan^{-1}
(\sqrt{|s_m/s_b|})$, then $B$ can not 
become stably established within the patch, as the local adaptation prevented by
migrational swamping \citep[see also][ for a review]{Lenormand}.
 \citet{barton1987establishment}, again working in one dimension,
 showed that this critical patch size also held in a stochastic
 framework \citep[see also the work of][]{Polk}. 
For patches above this size, \citet{barton1987establishment}
 founds that mutations appearing further than a few multiples of
 $\sigma/\sqrt{2|s_m|}$ away from such a patch have vanishingly small
 probability of becoming established within the patch. 
Barton also showed that mutations appearing within the patch have probability 
strictly less but approaching that $2s_b$ of establishment as the size
of the patch increased. Thus for large enough patches the probability
of establishment of a locally favorable allele approaches the standard
establishment probability ($2s_b$) for a panamitic population
\citep[][]{haldane, fisher}, a result that also (generally) holds for
a geographically spread population experiencing a uniform selection
pressure \citep{Maruyama, cherry}.

%Maruyama: On the fixation probability of mutant genes in a subdivided population*

@@

\subsubsection{Mutational influx}
\label{ss:patchymutation}
Consider first a single, isolated poor habitat patch of area $A$ in
which no $B$ allele has yet become established. As
we are interested in local adaptation that can occur we will assume
that our patch is considerably larger than the cutoff for local
establishment, i.e. $\sqrt{A} \gg 2 \tan^{-1} (\sqrt{|s_m/s_b|})$.

Let $p(x)$ be the probability that a new allele arising at location
$x$ away from the patch fixes within the poor habitat patch.
The function $p(x)$ can be found in terms of Jacobi elliptic functions,
XXX as shown in the Appendix, but the expressions are particularly unhelpful.
The total successful mutational influx per generation is then given by $\int \rho \mu p(x) dx$,
and depends in a complicated way on the relationship of patch width and selection coefficients,
but still scales linearly with the population mutational influx $\rho \mu$.
Therefore, we define $p_f = \int p(x) dx$.  If the width of the patch, $\sqrt{A}$, is large
relative to $\sigma/\sqrt{2s_m}$ then a reasonable approximation is to
set $p_f \approx 2 s_b A$. This follows from the fact that if the
patch is large than the majority of the mutations that could become
established within the patch, arose within the patch.

The rate at which mutations arise and colonize the patch is on the order of $\rho \mu
p_f$ generations, which will approximately be $2 s_b \rho \mu$. If
this rate is low,  then the waiting time till a mutation arises that
will become locally established within the patch is exponentially
distributed with mean $2 s_b \rho \mu$.  Assuming that once a mutation
becomes established
 it quickly reaches its equilibrium frequency across the patch, the
 expected time it takes for our patch to become colonized by the $B$
 allele is $1/(2 s_b \rho \mu)$.

\subsubsection{Migration rates}
\label{ss:patchymigration}

Now suppose that there are two patches, each of area $A$, of the poor
habitat type separated by distance $R$. Consider the case where the
allele has arisen and become established in the first patch, but has not yet appeared in the second.
As above, any copy of the new allele that appears at location $x$ relative to the second, unadapted patch 
has probability approximately $p(x)$ of fixing locally in this patch,
whether the allele is a new mutation or is descended from the occupants of the first patch.
Due to mutation--selection balance there will be a small number of alleles descended from those of the first patch (``migrants'') 
present in the regions where they are disadvantageous;
the equilibrium frequency may again be expressed in terms of Jacobi elliptic functions (Appendix XXX, citeXXX),
although the equations are again not particularly helpful.
However, we do know that if $q(r)$ is the expected proportion present at mutation--selection balance at distance $r$ from the first patch
in the absence of the second patch, that for large $r$ the proportion decays exponentially, $q(r) \approx C \exp( -|\sqrt{s_m} r / \sigma|)$,
where $C$ is a constant depending on the structure of the populations and the selection coefficients.

Combining this with the calculations above allows us to get a rough sense of the scale of migration between separated patches,
supposing that patches are large enough for local fixation to be
stable. If the alleles were not advantageous in the second patch,
the average numbers of mutant alleles present near the second patch would be around $\rho A \exp( - |\sqrt{s_m} R/\sigma| )$;
so the waiting time until some migrant allele fixes in the second patch would be something like $T_m = ( 2 s_b \rho A \exp( - |\sqrt{s_m} R/\sigma|)^{-1}$.
If $T_m$ is large, then we would expect successful migrants to be rare,
so that if the second patch adapts through migration, it is likely to come from only a single migrant.
We leave aside the converse situation, for which there is likely to be more than one migrant
that founds the population mutant individuals in the new patch.

\subsection{Patchy Selection -- Results} 
\label{ss:discreteresults}

We will assume that once a mutation escapes loss and becomes
established in the poor habitat patch it reaches its equilbrium frequency very rapidly.

If environment is continuous, with separated patches where the new allele is advantageous,
we can only get a rough sense of whether the new adaptation is likely to spread through mutation (i.e.\ in parallel)
or through migration.
Taking the expressions from sections \ref{ss:patchymutation} and \ref{ss:patchymigration} above,
we see that if patches each have sufficiently large area $A$,
then the mutational input in each is on the order of $2 \rho \mu A$,
while the influx due to migration between patches separated by distance $R$ is on the order $2 \rho A \exp(- |\sqrt{s_m} R/\sigma|)$. 
If the input from mutation and migration are comparable, i.e.\ if $\mu \approx \exp( -|\sqrt{s_m} R/\sigma| )$,
then adaptation is expected to take place both by migration and mutation.
If we take $\mu = 10^{-8}$, then this is satisfied if $R/\sigma \approx 18/\sqrt{s_m}$.
so if the disadvantage to the new mutation between patches is fairly strong (say, $s_m=.05$),
then the distance between patches must be about a hundred times the dispersal distance (i.e.\ $R/\sigma \approx 100$);
while if $s_m$ is much smaller, say $s_m = .001$, 
then the patches must be separated by many hundreds of times the dispersal distance (i.e.\ $R/\sigma \approx 500$).
The situation does not change much if $\mu$ is larger -- even if $\mu = 10^{-6}$, 
the required separation between patches is only reduced by 60\%.
If the separation between patches is much larger -- say, by a factor of ten -- 
then patches are effectively isolated -- most adaptation will be independent between patches.
On the other hand, if the separation is much smaller, then a single mutation is likely to fix everywhere.

@@

\begin{figure}[ht]
 \begin{center}
   \includegraphics{discrete-by-ratio}
 \caption{ %
 The mean number of types and the mean number of demes occupied by a sampled type, in a discrete model with 20 demes,
 plotted as a function of the ratio of migration to mutation probabilities.
 Note the log scale on the horizontal axis.
 \label{fig:discbyratio}
 }
 \end{center}
\end{figure}

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{natbib}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{color}
\usepackage{hyperref}
\usepackage{latexml} % for \iflatexml

@@

\newcommand{\gc}[1]{{ \it \color{green} (#1) } }

@@

Convergent adaptation can also occur within species, with individuals within a species 
adapted to the same environment in parallel through different large effect changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes REFs.
All else being equal the evolution of convergent phenotypes is more
likely with a 
higher mutational input, i.e. when the mutation target and population sizes are larger \citep{}. 
Standing variation for a trait can also increase the probability of convergence, 
as multiple pre-existing alleles can spread from low frequency, 
if our now adaptive alleles were previously not too deleterious \citep{Orr,Hermission}. 
The geographic distribution of populations can also shape the probability of parallel mutation within a species.
A geographically wide spread species is more likely to adapt by multiple mutations im parallel if dispersal 
is geographically limited, as populations will adapt via new mutations rather than waiting for migration \citep{RalphCoop}.

%When multiple rare changes sweep up from low frequency in parallel within a population this has been termed a soft sweep.\\

Intuitively, convergence is also more likely when geographically
separated populations adapt to ecologically similar conditions. The
probability of convergent adaptation rather than populations adapting
through allelic variants shared by migration, will be enhanced if
these adaptive alleles are maladapted in the environmentals separating
the populations. Such adverse conditions can strongly spatially restrict the spread of locally adapted alleles \citep{slatkin} 
and so might encourage local adaptation by parallel adaptation rather than migration.\\

@@

%In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
%where a continuous population was subjected to a selection pressure
%that was constant across the entire environment.

%and that a single mutational change was sufficient to adapt populations to the novel environment. 
%We assumed that there was no standing variation for the adaptive allele, such that parallel mutation must be due to multiple mutations after the environmental shift. In that setting we found that there was a characteristic geographic scale, over which we would expect multiple instances of our adaptive allele to have arisen in parallel. This characteristic length could be expressed in terms of a simple compound parameter determined by our parameters of interest.
%Recall questions of origin of adaptations

%References on parallel adapation, standing variation, etc.:
 % coop,
 % pennings and hermisson.
%Note pennings and hermisson found standing deleterious variation was very important.

%recall peromyscus examples with patchy environments

%Summarize previous paper

%Question assumptions of no standing variation and homogenous selection

@@

It takes only a single mutational change to create an allele ($B$) that adapts an individual to the
poor habitat type. This mutational change occurs at a low rate of
$\mu$ per chromosome per generation. We assume that this initially rare new
allele has fitness $1+s_B$ relative to the unmutated type ($b$) in the ``poor'' habitat patches,
and assume that it has fitness $1+s_m$ when rare in the intervening areas, with
$s_m<0$. To simplify matters we assume that the disadvantage $s_m$ ensures that on the relevant timescale,
the allele is very unlikely to fix locally in the regions where it is at a disadvantage.

@@

\paragraph{Previous work} We will make use of a number previous
results from the literature. \citet{slatkin1973geneflow}, working in
one dimensional deterministic framework, showed that if the
physical width of the patch is less than $2 \tan^{-1}
(\sqrt{|s_m/s_B|})$, then $B$ can not 
become stably established within the patch, as the local adaptation prevented by
migrational swamping \citep[see also][ for a review]{Lenormand}.
 \citet{barton1987establishment}, again working in one dimension,
 showed that this critical patch size also held in a stochastic
 framework \citep[see also the work of][]{Polk}. 
For patches above this size, \citet{barton1987establishment}
 founds that mutations appearing further than a few multiples of
 $\sigma/\sqrt{2|s_m|}$ away from such a patch have vanishingly small
 probability of becoming established within the patch. 
Barton also showed that mutations appearing within the patch have probability 
strictly less but approaching that $2s_B$ of establishment as the size
of the patch increased. Thus for large enough patches the probability
of establishment of a locally favorable allele approaches the standard
establishment probability ($2s_B$) for a panamitic population
\citep[][]{haldane, fisher}, a result that also (generally) holds for
a geographically spread population experiencing a uniform selection
pressure \citep{Maruyama, cherry}.

%Maruyama: On the fixation probability of mutant genes in a subdivided population*
%should we cite Cherry http://www.ncbi.nlm.nih.gov/pmc/articles/PMC1462507/

@@

\subsubsection{Time to the establishment of a locally adaptive allele
  due to mutational influx}
\label{ss:patchymutation}
Consider first a single, isolated poor habitat patch of area $A$ in
which no $B$ allele has yet become established. As
we are interested in local adaptation that can occur we will assume
that our patch is considerably larger than the cutoff for local
establishment, i.e. $\sqrt{A} \gg 2 \tan^{-1} (\sqrt{|s_m/s_B|})$.

Let $p(x)$ be the probability that a new allele arising at location
$x$ away from the patch fixes within the poor habitat patch.
The function $p(x)$ can be found in terms of Jacobi elliptic functions,
XXX as shown in the Appendix, but the expressions are particularly unhelpful.
The total successful mutational influx per generation is then given by $\int \rho \mu p(x) dx$,
and depends in a complicated way on the relationship of patch width and selection coefficients,
but still scales linearly with the population mutational influx $\rho \mu$.
Therefore, we define $p_f = \int p(x) dx$.  If the width of the patch, $\sqrt{A}$, is large
relative to $\sigma/\sqrt{2s_m}$ then a reasonable approximation is to
set $p_f \approx 2 s_B A$. This follows from the fact that if the
patch is large than the majority of the mutations that could become
established within the patch, arose within the patch.

The rate at which mutations arise and colonize the patch is
\begin{equation}
\lambda_{Mut} = \rho \mu p_f \approx 2 s_B \rho \mu.
\end{equation}

If this rate is low,  then the waiting time till a mutation arises that
will become locally established within the patch is exponentially
distributed with rate $\lambda_{Mut}$.  Assuming that once a mutation
becomes established
 it quickly reaches its equilibrium frequency across the patch, the
 expected time it takes for our patch to become colonized by the $B$
 allele is $T_{Mut} \approx 1/ \lambda_{Mut} $.

\subsubsection{Time to the establishment of a locally adaptive allele
  due to migrational influx}
\label{ss:patchymigration}
Now suppose that there are two patches, each of area $A$, of the poor
habitat type separated by distance $R$. Consider the case where the
$B$ allele has arisen and become established in the first patch, but has not yet appeared in the second.
As above, any copy of the new allele that appears at location $x$ relative to the second, unadapted patch 
has probability approximately $p(x)$ of fixing locally in this patch,
whether the allele is a new mutation or is descended from the occupants of the first patch.
Due to mutation--selection balance there will be a small number of $B$ alleles descended from those of the first patch (``migrants'') 
present in the regions where they are disadvantageous;
the equilibrium frequency may again be expressed in terms of Jacobi elliptic functions (Appendix XXX, citeXXX),
although the equations are again not particularly helpful.
However, we do know that if $q(r)$ is the expected proportion present at \gc{migration}--selection balance at distance $r$ from the first patch
in the absence of the second patch, that for large $r$ the proportion decays exponentially, $q(r) \approx C \exp( -|\sqrt{s_m} r / \sigma|)$,
where $C$ is a constant depending on the structure of the populations
and the selection coefficients. \gc{comment on size of C? Should we
  also use this approx. to get the approx. rate of establishing mutations
  arising outside of the patch, above. Feel like in the mutation rate
  part we could pull $s_B$ out front of all rates.}

\gc{Below are we setting $C=1$?}

Combining this with the calculations above allows us to get a rough sense of the scale of migration between separated patches,
supposing that patches are large enough for local fixation to be stable. If the alleles were not advantageous in the second patch,
the average number of mutant alleles present in a small area $\Delta$
near the second patch, due
to migration from the first patch, would be around $\rho A \exp( -
|\sqrt{s_m} R/\sigma| ) \Delta$. Assuming that the patch is
sufficiently large to give an allele a $2s_B$ probability of becoming established, the rate at which migrants find
themselves in the patch and become established  is approximately
\begin{equation}
\lamda_{Mig,R} = \rho A^2 s_B \exp( -|\sqrt{s_m} R/\sigma| ) \Delta
\end{equation}
. \gc{check my squared here. Do we want to use $q(R)$ here?|
If the establishment rate is low, the time it takes
a mutation to arise that will become established will be  $\sim Exp(\lambda_{mig,R})$

If  the time it takes for a mutation that it destined to 
become established in the second patch to become establish is short,
then  the expected waiting time until some migrant allele fixes in the
second patch will be $T_{Mig} = \lamda_{Mig,R} $.

\gc{Note that we can set patches sizes to be different}

\gc{think we've covered this next part by $q(r)$ being number of migs. such
that $2sq(r)$ is (for small $s$ the right answer.}
If $T_m$ is large, then we would expect successful migrants to be rare,
so that if the second patch adapts through migration, it is likely to come from only a single migrant.
We leave aside the converse situation, for which there is likely to be more than one migrant
that founds the population mutant individuals in the new patch.

\subsubsection{The probability of parallel adaptation between patches} 
\label{ss:probparallel}

As only a single mutation is required to adapt individuals to the
poor habitat patch, subsequent mutations that arise after an allele
becomes established at its equilibrium frequency on the patch gain no selective
benefit. Similarly an allele introduced into a patch by migration will
not be favored by selection to spread, if the patch has already been
colonized. Thus, mutations selective exclude each other from
patches, over short-time scales, and they will only slowly filter
together over longer time scales by drift.

We will assume that once a introduced $B$ allele (from migration or mutation) becomes
established in the poor habitat patch it reaches its equilbrium
frequency very rapidly. Assuming that we being from our geographically 
spread population containing no copies of the $B$ allele,

Under this approximation, and our selective
exclusion assumption, the second patch will become colonized by migration or
mutation, but not by both. This means that the question of whether the
second patch adapt in parallel, or through a shared migrant allele,
simply comes down to whether the a new mutation or a migrant allele
becomes established first.

If environment is continuous, with separated patches where the new allele is advantageous,
we can only get a rough sense of whether the new adaptation is likely to spread through mutation (i.e.\ in parallel)
or through migration.
Taking the expressions from sections \ref{ss:patchymutation} and \ref{ss:patchymigration} above,
we see that if patches each have sufficiently large area $A$,
then the mutational input in each is on the order of $2 \rho \mu A$,
while the influx due to migration between patches separated by distance $R$ is on the order $2 \rho A \exp(- |\sqrt{s_m} R/\sigma|)$. 
If the input from mutation and migration are comparable, i.e.\ if $\mu \approx \exp( -|\sqrt{s_m} R/\sigma| )$,
then adaptation is expected to take place both by migration and
mutation.

If we take $\mu = 10^{-8}$, then this is satisfied if $R/\sigma \approx 18/\sqrt{s_m}$.
so if the disadvantage to the new mutation between patches is fairly strong (say, $s_m=.05$),
then the distance between patches must be about a hundred times the dispersal distance (i.e.\ $R/\sigma \approx 100$);
while if $s_m$ is much smaller, say $s_m = .001$, 
then the patches must be separated by many hundreds of times the dispersal distance (i.e.\ $R/\sigma \approx 500$).
The situation does not change much if $\mu$ is larger -- even if $\mu = 10^{-6}$, 
the required separation between patches is only reduced by 60\%.
If the separation between patches is much larger -- say, by a factor of ten -- 
then patches are effectively isolated -- most adaptation will be independent between patches.
On the other hand, if the separation is much smaller, then a single mutation is likely to fix everywhere.

@@

%\begin{figure}[ht]
% \begin{center}
 %  \includegraphics{discrete-by-ratio}
 %\caption{ %
 %The mean number of types and the mean number of demes occupied by a sampled type, in a discrete model with 20 demes,
 %plotted as a function of the ratio of migration to mutation probabilities.
% Note the log scale on the horizontal axis.
% \label{fig:discbyratio}
 %}
 %\end{center}
%\end{figure}
